Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-11-19 manuscript/manuscript.tex
quick comments to intro and data sections

\doublespacing
\begin{document}

@@

 \tableofcontents % Optional: Table of Contents
 \listoffigures % List of Figures
 \listoftables % List of Tables

@@

\begin{abstract}
In the early 21st century, the open data movement began to transform 
societies and governments by promoting transparency,
innovation, and public engagement. The City of New York (NYC) has been at
the forefront of this movement since the enactment of the Open 
Data Law in 2012, leading to the creation of the NYC Open Data
portal. This portal currently hosts 2,700 datasets,
serving as a crucial resource for research across various domains, 
including health, urban development, and transportation. The 
success of this initiative highlights the importance of data 
curation in ensuring the utility and reliability of open datasets.

This paper examines the challenges of open data curation through a
case study of the NYC 311 Service Request (SR) dataset, addressing issues 
of data validity, consistency, and curation efficiency. Based on 
insights from this case study, we propose a set of data curation 
principles tailored for government\mbox{-}released open data. These principles 
aim to enhance data management practices and ensure 
the ongoing utility of open data. The paper concludes with 
actionable recommendations for enhancing data curation and outlines
general principles for the effective release of open data.

\end{abstract}

\begin{keywords}
  \kwd{Data cleansing}
  \kwd{Data Curation}
  \kwd{Data science}
  \kwd{NYC Open Data}
  \kwd{Open data}
  \kwd{Smart Cities}
  \kwd{Transparency}
\end{keywords}

@@

Research into data curation has explored these challenges in\mbox{-}depth. 
Among the earliest discussions, \citet{witt2009constructing} focused 
on developing data curation profiles tailored to specific contexts, 
setting a precedent for targeted data management strategies. 
Addressing broader challenges in data sharing and management, 
\citet{borgman2012conundrum} highlighted the complexities of 
research data distribution, emphasizing the need for robust 
strategies. This is complemented by \citet{hart2016ten}, who outlined 
essential principles for effective data management, particularly 
emphasizing meticulous curation practices. The utility of curated 
open data is vividly illustrated in public health and global challenges, 
where \citet{cantor2018facets} demonstrated the utility of curated 
data in evaluating community health determinants, and 
\citet{shankar2021data} observed its critical role during the 
COVID\mbox{-}19 pandemic in managing collective responses.

The contributions of this paper are twofold. First, we delve into
the specifics of data curation challenges using the NYC 311 Service
Request (SR) Data as a case study. This dataset serves as a prime 
example for examining key issues in data curation, including data 
validity, consistency, and curation efficiency. We illustrate these 
points with live examples drawn from our 
processing of the 311 SR data. Secondly, building upon insights 
gained from this case study, we propose a set of data curation 
principles tailored for government\mbox{-}released open data. These 
principles are designed to address the unique challenges 
and requirements observed in the curation of such datasets.

The paper is organized as follows. Section~\ref{sec:data} provides 
an overview of the history of the NYC 311 SR system and presents 
a summary of SR counts over a 2\mbox{-}year period from 2022 to 2023. 
Section~\ref{sec:issues} delves into specific data cleansing 
challenges affecting data quality and curation efficiency, 
including structural problems, adherence to the data dictionary, 
and the presence of missing, blank, or N/A entries. This section
 also investigates data field compliance with reference or acceptable values, 
highlights logical inconsistencies, and examines some concerning patterns 
in the data. We explore the balance between precision and accuracy 
and identify duplicate or redundant fields, along with observations 
on the \href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary}.

Section~\ref{sec:recommendations} offers 
practical recommendations for mitigating or resolving these issues, 
while Section~\ref{sec:discussion} encapsulates key insights and 
discusses the broader implications of our findings.

\section{NYC 311 Service Request (SR) Data} 
\label{sec:data}
The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non\mbox{-}emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non\mbox{-}emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone\mbox{-}based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile app channels in 2009, a
record high of 348,463 monthly service requests in August 2020 associated 
with the the COVID\mbox{-}19 pandemic, the 2021 expansion 
to include the MTA's city subway system (the biggest 
expansion in 311  history), and a record 3.23 million 
service requests in 2023.

@@

The impact of NYC 311 data extends beyond operational efficiency; it
has become instrumental in shaping City governance and community
engagement. Open data not only ensures governmental transparency
but also empowers researchers, civic developers, and the general
public. The data has been pivotal in providing advice on shelters
during emergencies, handling inquiries during the COVID\mbox{-}19 pandemic,
enforcing standards between landlords and tenants, reallocating taxi
routes based on analyses by the Taxi and Limousine Commission (TLC),
and improving responsiveness across City Agencies.

@@

\paragraph{Resolution Action Update Date}
A similar issue occurs when an SR is modified (and the \texttt{resolution\_action\_update\_date} is altered) well 
after the \texttt{closed\_date}.  When an SR is 
updated, the 311 software automatically populates the \texttt{resolution\_action\_update\_date}. While it is certainly 
possible, and even routine, to update an SR after it is closed, 
some of these updates seem to be well beyond any 
normal or expected timeframe, with some ``updates'' 
occurring over 2 years later. As shown in Figure~\ref{fig:resolution-violin}, 
there are 10,265 SRs that were updated >30 days after the 
\texttt{closed\_date} (Not that Figure~\ref(fig:resolution-violin) excludes 
values >730 days, an arbitrary cutoff to exclude 
infeasible dates such as \textit{1900-01-01}. The chart 
highlights the distribution of these post-closure 
 updates, with a notable concentration of updates occurring within 
the 30\mbox{-}to\mbox{-}200 day range. This pattern raises 
questions about whether such delayed updates are standard 
operational practice or indicative of a potential issue 
which warrants further investigation. It should be noted
that 84\% of these late updates are associated with the TLC agency
and 12\% with DSNY.

@@

\subsection{Accuracy and Precision:}
\label{sec:precision}
An question of precision vs. accuracy arises with the \texttt{latitude} 
and \texttt{longitude} fields. as well as with the duplicative 
\texttt{location} field. These fields are expressed as 
a 14\mbox{-}decimal number, e.g. \texttt{latitude} 40.86769186022511. 
Given that 1 degree of latitude at the equator is equal to 111.04 
kilometers, the ``1'' at the end of the decimal number represents 
approximately 1.1 nanometers (1/1,000,000,000 of a meter). (For 
reference a DNA molecule is approximately 2 nanometers in width.) The 
representation of the \texttt{latitude} and \texttt{longitude} fields
appears to be a classic case of extreme precision, but limited 
accuracy. All the \texttt{latitude}, \texttt{longitude}, and \texttt{location}
fields display this characteristic.

>>>>

\begin{document}

% \doublespacing

% \tableofcontents % Optional: Table of Contents
% \listoffigures % List of Figures
% \listoftables % List of Tables

@@

@@

\begin{abstract}
In the early 21st century, the open data movement began to transform 
societies and governments by promoting transparency,
innovation, and public engagement. The City of New York (NYC) has been at
the forefront of this movement since the enactment of the Open 
Data Law in 2012, leading to the creation of the NYC Open Data
portal. This portal currently hosts 2,700 datasets,
serving as a crucial resource for research across various domains, 
including health, urban development, and transportation. While the
open data movement has made vase amounts of data accessible, the
challenge of ensuring data quality and usability through effective
curation remain insufficiently addressed in the literature..
This paper examines these challenges through a
case study of the NYC 311 Service Request dataset, focusing on issues 
of data validity, consistency, and curation efficiency. Based on 
insights from this case study, we propose a set of data curation 
principles tailored for government-released open data aiming
to enhance data management practices and ensure 
the ongoing utility of open data.
\end{abstract}

\begin{keywords} % no repeating those in title
  \kwd{Data cleansing}
  \kwd{Data science}
  \kwd{NYC Open Data}
  \kwd{Smart Cities}
  \kwd{Transparency}
\end{keywords}

@@

Research into data curation has explored these challenges in\mbox{-}depth. 
Among the earliest discussions, \citet{witt2009constructing} focused 
on developing data curation profiles tailored to specific contexts, 
setting a precedent for targeted data management strategies. 
Addressing broader challenges in data sharing and management, 
\citet{borgman2012conundrum} highlighted the complexities of 
research data distribution, emphasizing the need for robust 
strategies. This is complemented by \citet{hart2016ten}, who outlined 
essential principles for effective data management, particularly 
emphasizing meticulous curation practices. The utility of curated 
open data is vividly illustrated in public health and global challenges, 
where \citet{cantor2018facets} demonstrated the utility of curated 
data in evaluating community health determinants, and 
\citet{shankar2021data} observed its critical role during the 
COVID-19 pandemic in managing collective responses.
\jy{The gap in the literature needs to be further highlighted to set 
  the stage for our contributions.}

The contributions of this paper are twofold. First, we delve into
the specifics of data curation challenges using the NYC 311 Service
Request (SR) Data as a case study. This dataset serves as a prime 
example for examining key issues in data curation, including data 
validity, consistency, and curation efficiency. We illustrate these 
points with live examples drawn from our 
processing of the 311 SR data. Secondly, building upon insights 
gained from this case study, we propose a set of data curation 
principles tailored for government-released open data. These 
principles are designed to address the unique challenges 
and requirements observed in the curation of such datasets.

The paper is organized as follows. Section~\ref{sec:data} provides 
an overview of the history of the NYC 311 SR system and presents 
a summary of SR counts over a 2\mbox{-}year period from 2022 to 2023. 
Section~\ref{sec:issues} delves into specific data cleansing 
challenges affecting data quality and curation efficiency, 
including structural problems, adherence to the data dictionary, 
and the presence of missing, blank, or N/A entries. This section
 also investigates data field compliance with reference or acceptable values, 
highlights logical inconsistencies, and examines some concerning patterns 
in the data. We explore the balance between precision and accuracy 
and identify duplicate or redundant fields, along with observations 
on the \href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary}.
Section~\ref{sec:recommendations} offers 
practical recommendations for mitigating or resolving these issues, 
while Section~\ref{sec:discussion} encapsulates key insights and 
discusses the broader implications of our findings.

\section{NYC 311 Service Request (SR) Data} 
\label{sec:data}

The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non\mbox{-}emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non\mbox{-}emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone\mbox{-}based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile app channels in 2009, a
record high of 348,463 monthly service requests in August 2020 associated 
with the the COVID\mbox{-}19 pandemic, the 2021 expansion 
\jy{use full name of MTA because it is used only once.}
to include the MTA's city subway system (the biggest 
expansion in 311  history), and a record 3.23 million 
service requests in 2023.

@@

The impact of NYC 311 data extends beyond operational efficiency; it
has become instrumental in shaping City governance and community
engagement. Open data not only ensures governmental transparency
but also empowers researchers, civic developers, and the general
public. The data has been pivotal in providing advice on shelters
during emergencies, handling inquiries during the COVID\mbox{-}19 pandemic,
enforcing standards between landlords and tenants, reallocating taxi
routes based on analyses by the Taxi and Limousine Commission (TLC),
and improving responsiveness across City Agencies.
\jy{need to expand the importance of 311 SR data to the research
  community: such as
  resource allocation/response \citep{zha2014profiling, raj2021swift},
  social equity \citep{kontokosta2021bias},
  street flood \citep{agonafir2022understanding}, etc.}

@@

\paragraph{Resolution Action Update Date}
A similar issue occurs when an SR is modified (and the \texttt{resolution\_action\_update\_date} is altered) well 
after the \texttt{closed\_date}.  When an SR is 
updated, the 311 software automatically populates the \texttt{resolution\_action\_update\_date}. While it is certainly 
possible, and even routine, to update an SR after it is closed, 
some of these updates seem to be well beyond any 
normal or expected timeframe, with some ``updates'' 
occurring over 2 years later. As shown in Figure~\ref{fig:resolution-violin}, 
there are 10,265 SRs that were updated >30 days after the 
\texttt{closed\_date} (Not that Figure~\ref{fig:resolution-violin} excludes 
values >730 days, an arbitrary cutoff to exclude 
infeasible dates such as \textit{1900-01-01}. The chart 
highlights the distribution of these post-closure 
 updates, with a notable concentration of updates occurring within 
the 30--200 day range. This pattern raises 
questions about whether such delayed updates are standard 
operational practice or indicative of a potential issue 
which warrants further investigation. It should be noted
that 84\% of these late updates are associated with the TLC agency
and 12\% with DSNY.

@@

\subsection{Accuracy and Precision:}
\label{sec:precision}
An question of precision vs. accuracy arises with the \texttt{latitude} 
and \texttt{longitude} fields. as well as with the duplicative 
\texttt{location} field. These fields are expressed as
\jy{This is a artifact of R reading them in as doubles. Single
  precision should be sufficient as illustrated in my Chance manuscript.}
a 14\mbox{-}decimal number, e.g. \texttt{latitude} 40.86769186022511. 
Given that 1 degree of latitude at the equator is equal to 111.04 
kilometers, the ``1'' at the end of the decimal number represents 
approximately 1.1 nanometers (1/1,000,000,000 of a meter). (For 
reference a DNA molecule is approximately 2 nanometers in width.) The 
representation of the \texttt{latitude} and \texttt{longitude} fields
appears to be a classic case of extreme precision, but limited 
accuracy. All the \texttt{latitude}, \texttt{longitude}, and \texttt{location}
fields display this characteristic.
